Megan's code for Isoviz

### 2023-04-04 temp_code/MS_isoviz_code.Rmd
Updated MS_isoviz_code

# 230327
Write a function that maps leafcutter cluster data back to isoforms 
- use psl/gtf, advantage of psl is it is compatible with long read generated gtf which lacks UTR info, it will likely become increasingly common to use long read data to determine which isoforms are present in a specific cell type
- input gene name and cell type, start with RBFOX2 and A375
- output is clusters with counts and linked isoforms
- print image filtered for cluster with <=4 junctions but print pdf with all
- if junctions doesn't map back, have a parameter to include/not include as novel
- go back and only plot isoforms with junction supporting data
- eventually also want to call Andrews predictions and get best predicted guide for junction
- could we even have a screen design setting? to scale it up?

@@

--includeconst
```{r}
# This is data from our A375 screen cells and is the most accurate for our work
# now this includes constitutive exons which have a single junction per cluster
# required 30 total reads per cluster and junction to be 10% of total
# total reads in a375-cyt_Aligned.q20.sort.bam: 43,407,715
clusters <- read_tsv("~/isoviz/data-raw/a375_screen_cells_perind.constcounts.gz", col_names = FALSE, skip = 1) # 59,783
clusters %<>% 
  separate(X1, into = c("coords", "counts"), sep = " ") %>% 
  separate(counts, into = c("junc.counts", "cluster.counts"), sep = "/", convert = TRUE) %>%
  separate(coords, into = c("chr", "start", "end", "name"), sep = ":", convert = TRUE) %>% 
  separate(name, into = c("same", "cluster.n", "extra"), sep = "_") %>% 
  select(-same, -extra)

@@

a375_screen_lfc_clusters = as.data.table(clusters)
save(a375_screen_lfc_clusters, file = "~/isoviz/data/a375_screen_lfc_clusters.rda")

@@

Function: Map junction to isoform
```{r}
load(file = "~/isoviz/data/iso_intron_data.rda")
load(file = "~/isoviz/data/a375_screen_lfc_clusters.rda")

# map junctions function- for now, it isn't using cell type but eventually we should change this
map_junction = function(gene = "ENSG00000001167.15", cell_type = "A375"){

@@

  # join iso data with leafcutter data and mark any unannotated junctions
  const_junc_list = a375_screen_lfc_clusters %>% 
    group_by(cluster.n) %>% 
    tally() %>% filter(n == 1) %>% ungroup()
  a375_screen_lfc_clusters %<>% mutate(junc.type = ifelse(cluster.n %in% const_junc_list$cluster.n, "Common", "Unique"))
  gene_cluster = gene_iso_data %>% 
    left_join(a375_screen_lfc_clusters, by = c("chr", "intron_starts" = "start", "intron_ends" = "end")) %>%
    filter(!is.na(cluster.n))

@@

  output = a375_screen_lfc_clusters %>% filter(cluster.n %in% cluster_ids) %>% group_by(cluster.n) %>% mutate(junc.per.cluster = n()) %>% 
    left_join(gene_iso_data, by = c("chr", "start" = "intron_starts", "end" = "intron_ends")) %>% 
    mutate(junc.usage = round((junc.counts/cluster.counts)*100, digits = 0), 
           transcript_name = ifelse(is.na(transcript_name), paste0(cluster.n, "_Unannotated"), transcript_name),
           trans_id = ifelse(is.na(trans_id), paste0(cluster.n, "_Unannotated"), trans_id),
           gene_id = ifelse(is.na(gene_id), gene, gene_id),
           gene_name = ifelse(is.na(gene_name), name, gene_name),
           strand = ifelse(is.na(strand), gene_strand, strand))

@@

```{r}
to_plot = map_junction("ENSG00000001167.15") # NFYA
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/nfya_isoforms.pdf", width = 8, height = 4)

to_plot = map_junction("ENSG00000100320") # RBFOX2, alt start is missing because it's intron doesnt overlap another and counts are only 8 so included in consitutive either
plot_junction_to_isoform(to_plot)
```

>>>>

# 230327
Write a function that maps leafcutter cluster data back to isoforms 
- use psl/gtf, advantage of psl is it is compatible with long read generated gtf which lacks UTR info, it will likely become increasingly common to use long read data to determine which isoforms are present in a specific cell type
- input gene name and cell type, start with RBFOX2 and A375
- output is clusters with counts and linked isoforms
- print image filtered for cluster with <=4 junctions but print pdf with all
- if junctions doesn't map back, have a parameter to include/not include as novel
- go back and only plot isoforms with junction supporting data
- eventually also want to call Andrews predictions and get best predicted guide for junction
- could we even have a screen design setting? to scale it up?
- also, can use the code and plots to quantify knockdown
- need to control for library size- tpm of gene?

@@

--includeconst
```{r}
# This is data from our A375 screen cells and is the most accurate for our work
# now this includes constitutive exons which have a single junction per cluster
# required 30 total reads per cluster and junction to be 10% of total
clusters <- read_tsv("~/isoviz/data-raw/a375_screen_cells_perind.constcounts.gz", col_names = FALSE, skip = 1) # 59,783
clusters %<>% 
  separate(X1, into = c("coords", "counts"), sep = " ") %>% 
  separate(counts, into = c("junc.counts", "cluster.counts"), sep = "/", convert = TRUE) %>%
  separate(coords, into = c("chr", "start", "end", "name"), sep = ":", convert = TRUE) %>% 
  separate(name, into = c("same", "cluster.n", "extra"), sep = "_") %>% 
  select(-same, -extra)

@@

clusters_table = as.data.table(clusters)
save(clusters_table, file = "~/isoviz/data/A375_lfc_clusters.rda")

# HEK293 cells- our data
clusters <- read_tsv("~/isoviz/data-raw/HEK-NFYA-rep2-empty_perind.constcounts.gz", col_names = FALSE, skip = 1) # 46,102
clusters %<>% 
  separate(X1, into = c("coords", "counts"), sep = " ") %>% 
  separate(counts, into = c("junc.counts", "cluster.counts"), sep = "/", convert = TRUE) %>%
  separate(coords, into = c("chr", "start", "end", "name"), sep = ":", convert = TRUE) %>% 
  separate(name, into = c("same", "cluster.n", "extra"), sep = "_") %>% 
  select(-same, -extra)

# remove clusters with no counts
# filter out chrM and chrY clusters
clusters %<>% filter(cluster.counts > 0, chr != "chrY", chr != "chrM")
clusters_table = as.data.table(clusters)
save(clusters_table, file = "~/isoviz/data/HEK293_lfc_clusters.rda")

# 35738736-36028824
singletons <- read_tsv("~/knowles_lab/Megan/230320_cas13_validation/junctions/HEK-NFYA-rep2-empty_singletons_perind.constcounts.gz", col_names = FALSE, skip = 1)
singletons %<>% 
  separate(X1, into = c("coords", "counts"), sep = " ") %>% 
  separate(counts, into = c("junc.counts", "cluster.counts"), sep = "/", convert = TRUE) %>%
  separate(coords, into = c("chr", "start", "end", "name"), sep = ":", convert = TRUE) %>% 
  separate(name, into = c("same", "cluster.n", "extra"), sep = "_") %>% 
  select(-same, -extra)

singletons %>% filter(chr == "chr22", start >= 35738736, end <= 36028824)

# hESC- our data
clusters <- read_tsv("~/isoviz/data-raw/21c_hESC_RB2_empty_perind.constcounts.gz", col_names = FALSE, skip = 1)
clusters %<>% 
  separate(X1, into = c("coords", "counts"), sep = " ") %>% 
  separate(counts, into = c("junc.counts", "cluster.counts"), sep = "/", convert = TRUE) %>%
  separate(coords, into = c("chr", "start", "end", "name"), sep = ":", convert = TRUE) %>% 
  separate(name, into = c("same", "cluster.n", "extra"), sep = "_") %>% 
  select(-same, -extra)

# remove clusters with no counts
# filter out chrM and chrY clusters
clusters %<>% filter(cluster.counts > 0, chr != "chrY", chr != "chrM")
clusters_table = as.data.table(clusters)
save(clusters_table, file = "~/isoviz/data/hESC_lfc_clusters.rda")

@@

Function: Map junction to isoform
```{r}
load(file = "~/isoviz/data/iso_intron_data.rda")
#load(file = "~/isoviz/data/a375_screen_lfc_clusters.rda")
#load(file = "~/isoviz/data/HEK293_lfc_clusters.rda")

# is it better to load rda within a function?
# map junctions function- for now, it isn't using cell type but eventually we should change this
map_junction = function(gene = "ENSG00000001167.15", cell_type = "HEK293"){

@@

  # join iso data with leafcutter data and mark any unannotated junctions
  # consider cell type
  lfc_clusters = paste0(cell_type, "_lfc_clusters")
  load(file = paste0("~/isoviz/data/", lfc_clusters, ".rda"))
  const_junc_list = clusters_table %>% 
    group_by(cluster.n) %>% 
    tally() %>% filter(n == 1) %>% ungroup()
  clusters_table %<>% mutate(junc.type = ifelse(cluster.n %in% const_junc_list$cluster.n, "Common", "Unique"))
  gene_cluster = gene_iso_data %>% 
    left_join(clusters_table, by = c("chr", "intron_starts" = "start", "intron_ends" = "end")) %>%
    filter(!is.na(cluster.n))

@@

  output = clusters_table %>% filter(cluster.n %in% cluster_ids) %>% group_by(cluster.n) %>% mutate(junc.per.cluster = n()) %>% 
    left_join(gene_iso_data, by = c("chr", "start" = "intron_starts", "end" = "intron_ends")) %>% 
    mutate(junc.usage = round((junc.counts/cluster.counts)*100, digits = 0), 
           transcript_name = ifelse(is.na(transcript_name), paste0(cluster.n, "_Unannotated"), transcript_name),
           trans_id = ifelse(is.na(trans_id), paste0(cluster.n, "_Unannotated"), trans_id),
           gene_id = ifelse(is.na(gene_id), gene, gene_id),
           gene_name = ifelse(is.na(gene_name), name, gene_name),
           strand = ifelse(is.na(strand), gene_strand, strand), cell_line = paste0(cell_type))

@@

```{r}
# NFYA
to_plot = map_junction("ENSG00000001167.15", cell_type = "A375") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/nfya_isoforms.pdf", width = 8, height = 4)

to_plot = map_junction("ENSG00000001167.15", cell_type = "HEK293") 
plot_junction_to_isoform(to_plot)
# for height, count rows and divide by 2
ggsave("~/isoviz/plots/nfya_isoforms_hek293.pdf", width = 8, height = 2)

to_plot = map_junction("ENSG00000001167.15", cell_type = "hESC") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/nfya_isoforms_hESC.pdf", width = 8, height = 2)

# RBFOX2
to_plot = map_junction("ENSG00000100320", cell_type = "A375") # alt start is missing because it's intron doesnt overlap another and counts are only 8 so included in consitutive either
plot_junction_to_isoform(to_plot)

to_plot = map_junction("ENSG00000100320", cell_type = "HEK293") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/rbfox2_isoforms_hek293.pdf", width = 8, height = 9)

to_plot = map_junction("ENSG00000100320", cell_type = "hESC") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/rbfox2_isoforms_hESC.pdf", width = 8, height = 8.5)

Generate a junction file of all possible junctions sequences for Andrew to make predictions
- use gencode psl
- want to make sure to have junc id that will correspond to leafcutter (can merge by sequence at first)
- can make this information into a table
```{r}
load(file = "~/isoviz/data/iso_intron_data.rda")
iso_intron_data %>% group_by(transcript_type) %>% tally()
iso_intron_data %>% group_by(chr) %>% tally()
iso_intron_data %>% filter(is.na(chr))

# filter for protein coding and lncRNAs
iso_intron_pc = iso_intron_data %>% filter(transcript_type == "protein_coding" | transcript_type == "lncRNA")

# remove transcript ids and collapse on junction coords
collapsed_junctions = iso_intron_pc %>% select(-trans_id, -transcript_name, -gene_type, -transcript_type) %>% 
  distinct() %>% select(chr, start = intron_starts, end = intron_ends, gene_id, gene_name, strand) # 281,880 junctions- how many show any uniqueness?

# to address how many annotated isoforms have a truly unique junction
# first filter the genes with more than one transcript
iso_intron_pc %>% select(gene_id, trans_id) %>% distinct() # 85,586 transcripts
n_df = iso_intron_pc %>% select(gene_id, trans_id) %>% distinct() %>% group_by(gene_id) %>% mutate(n_transcripts = n()) %>% ungroup() # 33,413
gene_list = n_df %>% select(-trans_id) %>% distinct() %>% filter(n_transcripts > 1) %>% select(gene_id) # 17,035 genes have more than 1 isoform

iso_pc_df = iso_intron_pc %>% full_join(n_df) %>% filter(gene_id %in% gene_list$gene_id) %>%
  group_by(chr, intron_starts, intron_ends, gene_id, strand, gene_name, n_transcripts) %>% mutate(n_trans_per_junc = n(), commonality = round((n_trans_per_junc/n_transcripts)*100, 0)) %>%
  mutate(junction_category = ifelse(n_trans_per_junc == 1, "fully_unique", ifelse(commonality == 100, "common", "partial_unique"))) %>% ungroup()

# junction level numbers- there are ~50,000 fully unique junctions for pc genes, meaning they belong to a singular isoform
iso_pc_df %>% select(chr, intron_starts, intron_ends, strand, gene_id, gene_name, gene_type, junction_category) %>% distinct() %>% 
  group_by(gene_type, junction_category) %>% tally()

# isoform-level, how many isoforms are uniquely targetable
iso_pc_df %>% filter(junction_category == "fully_unique") %>% select(gene_id, gene_name, trans_id, transcript_name, gene_type) %>% distinct() # 47,657 transcripts are uniquely targetable
iso_pc_df %>% filter(junction_category == "fully_unique") %>% select(gene_id, gene_name, trans_id, transcript_name, gene_type) %>% distinct() %>%
  distinct(gene_name) # 16,801 genes
